---
title: "The Relationship of Student Attendance and Student Proficiency in English, Language, and Arts "
author: "Tram Anh Hoang, Haitham Anbar, Maria Coronado, Erick Njue"
date: "`r Sys.Date()`"
output: pdf_document
bibliography: references.bib 
---
#### Comments from Brittany Spinner 
In your _YAML_ I didn't know that the code r Sys.Date(), with the tick marks to do inline code, would produce the current date everytime the document is rendered. That is so cool and I will have to do this in the future! 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, results='hide')


library(here)
library(rio)
library(readr)
library(janitor)
library(tidyverse)
library(knitr)

```

#### Comment 
Do you need both {rio} and {readr}? 


# Abstract

# Introduction
We use data from the Oregon Department of Education (ODE) to examine the following research questions:  

* Does attendance rate affect student proficiency in English, Language, and Arts? and;
* Does such a relationship vary for different racial and ethnic groups?  
 
We use data of student attendance rate  and student proficiency rate  from the ODE assessment group reports for all the schools in the state. We will base our assessment on the latest academic year available for both data sets, which is 2021 – 2022.

#### Comments 
These are excellent research questions, maybe label them as *RQ1* and *RQ2* (only if you want to) and add hypothesis to each research question. You'll want to provide more details about your sample such as age range, sample size = n, and race and ethnic groups, etc. This can come after the intro in a separate paragraph _Sample_. Good description of the dataset though and where it came from and the year. Maybe list the exact variable names you're looking at as well (?). 

# Methods
In this research, we run a regression model on attendance data and proficiency data of schools in Oregon. We obtain two different data sets from the ODE: attendance rate by school and student group, and proficiency in English, Language, and Arts by student group. 

## Data description and preparation

We used [@R] package [@here] to read in the two datasets from Oregon Department of Education (ODE)into the [@R] environment. We use [@janitor] to clean the names. After preparing the datasets, we display them as tables using [@tidyverse].

```{r read_data}

race <- import(here("data", "pagr_schools_ela_tot_raceethnicity_2122.csv")) %>% 
  clean_names() %>%
  as_tibble()
view(race)

attend <- import(here("data", "regularattenders_report_2122.csv")) %>% 
  clean_names() %>%
  as_tibble()
view(attend)
```

The first data set is student performance in English Language and Arts (ELA) from all schools in the state. The data set provides the number and percentage of students that are proficient in each of the four levels of ELA proficiency, stratified by school, race and ethnicity, and grade level. This data set has `r nrow(race)` rows and `r ncol(race)` columns.  

The second data set is student attendance data from all the schools in Oregon. This data set provides the number and percentage of regular attenders as well as those of chronically absent students, stratified by school and either race and ethnicity or grade level. This data set has `r nrow(attend)` rows and `r ncol(attend)` columns.  

We clean the two data sets by removing rows with no data and rows with suppressed data, marked as “- -” and “*” respectively. We use {dplyr} package functions in [@tidyverse] to do so. 

#### Comments 
Do the participants from each dataset have the same IDs? Meaning, does the attendance for, lets say, a student with the ID 123 match the ID for that student's performance scores in ELA dataset? In other words, are these the same exact kids that you are comparing attendance and performance with, and do they have matching IDs in each dataset? You'll have to get rid of all students from ELA that don't have attendance data in and vice versa. This is a kind of join, I think _inner_join_? _inner_join_ keeps only the data that matches in both datasets. I think this would be a good option here. Then proceed with filtering etc. below. 

```{r clean_attendance}
#Summarizing to check the existing student groups in the file
attendsum <- attend %>% 
 group_by(student_group) %>% 
  summarise()
attendsum


#Filtering the dataframe leaving only student groups related to race. 
attend1 <- attend %>% 
   filter(student_group == "American Indian/Alaska Native" |
          student_group == "Asian"|
          student_group == "Black/African American" |
          student_group == "Hispanic/Latino" |
          student_group == "Multi-Racial" |
          student_group == "Native Hawaiian/Pacific Islander" |
          student_group == "White") %>%
  mutate(student_group=recode(student_group,
                              "Native Hawaiian/Pacific Islander" = "Pacific Islander"))
# why did you rename native hawaiin/pacific islander to "pacific islander only"? Not sure I would do this, but if there is a specific reason for doing this, such as it helps with coding etc., I totally get it. 


#Removing rows without data
attend1 <- attend1%>% 
  filter(students_included != "*")
# this would perhaps be more beneficial if you combined the datasets using {inner_join} and had everything in once place (?). 
    
# Note: All schools with >95% and <5% regular attenders were eliminated. 
## Why? Does this skew the data in some way?
#These rows corresponded with schools where less than 9 students were included. 

dim(attend1)
view(attend1)
```

```{r clean_profifiency}
#Finding the races considered in the file
racesum <- race %>% 
 group_by(student_group) %>% 
  summarise()


#Filtering the dataframe leaving only student groups related to race.
#Filtered again to remove rows without data on proficiency data
#Transformed number of participants to numeric and the filtered to all >= than 10 students. 
race1 <- race %>% 
   select(-grade_level) %>% 
     filter(student_group == "American Indian/Alaskan Native" |
          student_group == "Asian"|
          student_group == "Black/African American" |
          student_group == "Hispanic/Latino" |
          student_group == "Multi-Racial" |
          student_group == "Pacific Islander" |
          student_group == "White") %>% 
 filter(number_proficient != "*" &
        number_proficient != "--") %>% 
 transform(number_of_participants = as.numeric(number_of_participants)) %>% 
 filter(number_of_participants >= 10)
race1

#Transformed all columns to numbers to be able to compute them. The file had them as characters.
race2<- race1 %>% 
  transform(
    number_proficient = as.numeric(number_proficient), 
    percent_proficient_level_3_or_4 = as.numeric(percent_proficient_level_3_or_4),
    number_level_4 = as.numeric(number_level_4), 
    percent_level_4 = as.numeric(percent_level_4), 
    number_level_3 = as.numeric(number_level_3), 
    percent_level_3 = as.numeric(percent_level_3), 
    number_level_2 = as.numeric(number_level_2), 
    percent_level_2 = as.numeric(percent_level_2), 
    number_level_1 = as.numeric(number_level_1),
    percent_level_1 = as.numeric(percent_level_1), 
    participation_rate = as.numeric(participation_rate)
    )
race2
view(race2)

#Grouped data by school_id and student_group, summed the number of participants and number of proficient students, and re-calculate the percentage of proficient students.
race3 <- race2 %>% 
 group_by(district_id,school_id, student_group) %>% 
 summarise(number_proficient = sum(number_proficient),
           number_of_participants=sum(number_of_participants)) %>%
  mutate(percent_proficient = number_proficient/number_of_participants*100)

race3

```

After cleaning, we merge the two data sets by matching schools’ IDs and student race and ethnicity. Rows with attendance data only or proficiency data only will be removed. We used the _left_join_ in the {dplyr} in [@tidyverse]

```{r join}
#Check that 2 data sets have matching student groups
unique(attend1$student_group)
unique(race3$student_group)

#Join
attend_race <- left_join(attend1, race3, by=c("district_id"="district_id", "institution_id"="school_id", "student_group"="student_group")) %>%
  drop_na()
## Ok, so you did all the cleaning first and then joined. I wonder if joining first is better? I have no idea though. and should you do inner_join instead of left-join, but I guess if you already got rid of all the missing data then maybe it isn't an issue? 


view(attend_race)
```
## Methods


# Results

```{r summary_data, results='markup'}

#Pivot wider by student_group to have columns by races:
prof_by_race <- attend_race %>%
  group_by(institution_type, student_group) %>% 
  summarise(number_of_participants = sum(number_of_participants),
            number_proficient = sum(number_proficient)) %>% 
  mutate(percent_prof_race = (number_proficient/number_of_participants*100)) %>% 
  mutate(percent_prof_race = round(percent_prof_race, 2)) %>% 
 pivot_wider(names_from = "student_group",
 values_from = "percent_prof_race") %>% 
  clean_names()

#Doing an additional pivot longer followed by a pivot wider to make the table look good:
prof_percent_by_type_of_inst <- prof_by_race %>% 
  select(-number_of_participants, -number_proficient) %>%
  pivot_longer( cols = 2:7,
                names_to = "race",
                values_to = "values") %>% 
  drop_na() %>% 
  pivot_wider(names_from = "race",
  values_from = "values") %>% 
  clean_names()

prof_percent_by_type_of_inst %>%
  kable(caption="Student Proficiency Rate by Race and Institution types",
        col.names=c("Institution type",
                    "Asian", 
                    "Black African American", 
                    "Hispanic/Latino",
                    "Multi racial",
                    "Pacific Islander",
                    "White"))

```
#### Comment 
What is the difference between school and high school? Does school mean grade school (i.e., K-8) and high school is 9-12th grade? 


```{r plot, results='asis'}

#Adding plot for attendance vs. proficiency
attend_race %>%
  ggplot(aes(x=percent_regular_attenders, y=percent_proficient)) +
  geom_point() +
  geom_smooth(method='lm') +
  theme_bw() +
  labs(title="Student Attendance and Proficiency Rate \n in English, Language, and Arts",
       subtitle="Data from Oregon Education Department",
       x="Rate of Regular Attendance (%)",
       y="Percentage of proficient students (%)",
       tag = "Plot 1")
# I wonder why the line for this figure isn't showing up in the graph, we should also see the x-axis scale (?). 
# I think this y label is wrong. you're not saying the percent of proficient students you're saying the score for proficiency in percent for students correct? 
# and since you got rid of >95% attendance and <5% attendance your x axis would be 5-95 and not 1-100. which i think it should be. I don't think you should use this constraint on attendance, but I would have to know why you are using it to say you shouldn't. BUT if you are going to keep this filter, I would say why and then make it known in the graph as well on the x axis. 

#Adding plot for attendance vs. proficiency, faceted by race
attend_race %>%
  ggplot(aes(x=percent_regular_attenders, y=percent_proficient,group=student_group)) +
  geom_point() +
  geom_smooth(method='lm') +  
  facet_wrap(~student_group) +
  theme_bw() + # I would do theme_minimal so you can see the lines of the graph. I tried this and the lines still didn't show so I'm not sure why they wont show up, but it would be nice to see the lines on the first graph at least. 
  labs(title="Student Attendance and Proficiency Rate \n in English, Language, and Arts by Different Races",
       subtitle="Data from Oregon Education Department",
       x="Rate of Regular Attendance (%)",
       y="Percentage of proficient students (%)",
       tag="Plot 2")

```



#### Comment 
These are super cool graphs! Well done! I think you should have someone run regression model(s) using lm() to see the significance of the slope and intercept. The graph representing the Asian population is interesting because it starts in the middle of the graph and the graph for the pacific islander population ends in the middle of the graph. Very unique! I love this facetwrap! 

#### Comment 
to run a regression model for the first graph you would use the following code:
```{r}
fit <- lm(formula = percent_regular_attenders ~ percent_proficient, data = attend_race)
summary(fit)
```


#### Comments 
Congrats! It looks like your results are significant! For every one unit (one percent in this case) increase in attendance, this graph represents a predicted .43% increase in proficiency scores. (I think... I'm not super familiar with these data though and I would review what I've done and cater it to your needs) 
Great job you guys! 

# Discussions

\newpage
# References


