---
title: "The Relationship of Student Attendance and Student Proficiency in English, Language, and Arts "
author: "Tram Anh Hoang, Haitham Anbar, Maria Coronado, Erick Njue"
date: "`r Sys.Date()`"
output: pdf_document
bibliography: references.bib 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, results='hide')


library(here)
library(rio)
library(readr)
library(janitor)
library(tidyverse)
library(knitr)

```


# Abstract

# Introduction
We use data from the Oregon Department of Education (ODE) to examine the following research questions:  

* Does attendance rate affect student proficiency in English, Language, and Arts? and;
* Does such a relationship vary for different racial and ethnic groups?  
 
We use data of student attendance rate  and student proficiency rate  from the ODE assessment group reports for all the schools in the state. We will base our assessment on the latest academic year available for both data sets, which is 2021 – 2022.

# Methods
In this research, we run a regression model on attendance data and proficiency data of schools in Oregon. We obtain two different data sets from the ODE: attendance rate by school and student group, and proficiency in English, Language, and Arts by student group. 

## Data description and preparation

We used [@R] package [@here] to read in the two datasets from Oregon Department of Education (ODE)into the [@R] environment. We use [@janitor] to clean the names. After preparing the datasets, we display them as tables using [@tidyverse].

```{r read_data}

race <- import(here("data", "pagr_schools_ela_tot_raceethnicity_2122.csv")) %>% 
  clean_names() %>%
  as_tibble()
view(race)

attend <- import(here("data", "regularattenders_report_2122.csv")) %>% 
  clean_names() %>%
  as_tibble()
view(attend)
```

The first data set is student performance in English Language and Arts (ELA) from all schools in the state. The data set provides the number and percentage of students that are proficient in each of the four levels of ELA proficiency, stratified by school, race and ethnicity, and grade level. This data set has `r nrow(race)` rows and `r ncol(race)` columns.  

The second data set is student attendance data from all the schools in Oregon. This data set provides the number and percentage of regular attenders as well as those of chronically absent students, stratified by school and either race and ethnicity or grade level. This data set has `r nrow(attend)` rows and `r ncol(attend)` columns.  

We clean the two data sets by removing rows with no data and rows with suppressed data, marked as “- -” and “*” respectively. We use {dplyr} package functions in [@tidyverse] to do so. 

```{r clean_attendance}
#Summarizing to check the existing student groups in the file
attendsum <- attend %>% 
 group_by(student_group) %>% 
  summarise()
attendsum


#Filtering the dataframe leaving only student groups related to race. 
attend1 <- attend %>% 
   filter(student_group == "American Indian/Alaska Native" |
          student_group == "Asian"|
          student_group == "Black/African American" |
          student_group == "Hispanic/Latino" |
          student_group == "Multi-Racial" |
          student_group == "Native Hawaiian/Pacific Islander" |
          student_group == "White") %>%
  mutate(student_group=recode(student_group,
                              "Native Hawaiian/Pacific Islander" = "Pacific Islander"))


#Removing rows without data
attend1 <- attend1%>% 
  filter(students_included != "*")
    
# Note: All schools with >95% and <5% regular attenders were eliminated. 
#These rows corresponded with schools where less than 9 students were included. 

dim(attend1)
view(attend1)
```

```{r clean_profifiency}
#Finding the races considered in the file
racesum <- race %>% 
 group_by(student_group) %>% 
  summarise()


#Filtering the dataframe leaving only student groups related to race.
#Filtered again to remove rows without data on proficiency data
#Transformed number of participants to numeric and the filtered to all >= than 10 students. 
race1 <- race %>% 
   select(-grade_level) %>% 
     filter(student_group == "American Indian/Alaskan Native" |
          student_group == "Asian"|
          student_group == "Black/African American" |
          student_group == "Hispanic/Latino" |
          student_group == "Multi-Racial" |
          student_group == "Pacific Islander" |
          student_group == "White") %>% 
 filter(number_proficient != "*" &
        number_proficient != "--") %>% 
 transform(number_of_participants = as.numeric(number_of_participants)) %>% 
 filter(number_of_participants >= 10)
race1

#Transformed all columns to numbers to be able to compute them. The file had them as characters.
race2<- race1 %>% 
  transform(
    number_proficient = as.numeric(number_proficient), 
    percent_proficient_level_3_or_4 = as.numeric(percent_proficient_level_3_or_4),
    number_level_4 = as.numeric(number_level_4), 
    percent_level_4 = as.numeric(percent_level_4), 
    number_level_3 = as.numeric(number_level_3), 
    percent_level_3 = as.numeric(percent_level_3), 
    number_level_2 = as.numeric(number_level_2), 
    percent_level_2 = as.numeric(percent_level_2), 
    number_level_1 = as.numeric(number_level_1),
    percent_level_1 = as.numeric(percent_level_1), 
    participation_rate = as.numeric(participation_rate)
    )
race2
view(race2)

#Grouped data by school_id and student_group, summed the number of participants and number of proficient students, and re-calculate the percentage of proficient students.
race3 <- race2 %>% 
 group_by(district_id,school_id, student_group) %>% 
 summarise(number_proficient = sum(number_proficient),
           number_of_participants=sum(number_of_participants)) %>%
  mutate(percent_proficient = number_proficient/number_of_participants*100)

race3

```

After cleaning, we merge the two data sets by matching schools’ IDs and student race and ethnicity. Rows with attendance data only or proficiency data only will be removed. We used the _left_join_ in the {dplyr} in [@tidyverse]

```{r join}
#Check that 2 data sets have matching student groups
unique(attend1$student_group)
unique(race3$student_group)

#Join
attend_race <- left_join(attend1, race3, by=c("district_id"="district_id", "institution_id"="school_id", "student_group"="student_group")) %>%
  drop_na()

view(attend_race)
```
## Methods

# Results

```{r summary_data, results='markup'}

#Pivot wider by student_group to have columns by races:
prof_by_race <- attend_race %>%
  group_by(institution_type, student_group) %>% 
  summarise(number_of_participants = sum(number_of_participants),
            number_proficient = sum(number_proficient)) %>% 
  mutate(percent_prof_race = (number_proficient/number_of_participants*100)) %>% 
  mutate(percent_prof_race = round(percent_prof_race, 2)) %>% 
 pivot_wider(names_from = "student_group",
 values_from = "percent_prof_race") %>% 
  clean_names()

#Doing an additional pivot longer followed by a pivot wider to make the table look good:
prof_percent_by_type_of_inst <- prof_by_race %>% 
  select(-number_of_participants, -number_proficient) %>%
  pivot_longer( cols = 2:7,
                names_to = "race",
                values_to = "values") %>% 
  drop_na() %>% 
  pivot_wider(names_from = "race",
  values_from = "values") %>% 
  clean_names()

prof_percent_by_type_of_inst %>%
  kable(caption="Student Proficiency Rate by Race and Institution types",
        col.names=c("Institution type",
                    "Asian", 
                    "Black African American", 
                    "Hispanic/Latino",
                    "Multi racial",
                    "Pacific Islander",
                    "White"))

```

```{r plot, results='asis'}

#Adding plot for attendance vs. proficiency
attend_race %>%
  ggplot(aes(x=percent_regular_attenders, y=percent_proficient)) +
  geom_point() +
  geom_smooth(method='lm') +
  theme_bw() +
  labs(title="Student Attendance and Proficiency Rate \n in English, Language, and Arts",
       subtitle="Data from Oregon Education Department",
       x="Rate of Regular Attendance (%)",
       y="Percentage of proficient students (%)",
       tag = "Plot 1")

#Adding plot for attendance vs. proficiency, faceted by race
attend_race %>%
  ggplot(aes(x=percent_regular_attenders, y=percent_proficient,group=student_group)) +
  geom_point() +
  geom_smooth(method='lm') +  
  facet_wrap(~student_group) +
  theme_bw() +
  labs(title="Student Attendance and Proficiency Rate \n in English, Language, and Arts by Different Races",
       subtitle="Data from Oregon Education Department",
       x="Rate of Regular Attendance (%)",
       y="Percentage of proficient students (%)",
       tag="Plot 2")

```

```{r regression}
attpro <- lm(formula = percent_regular_attenders~percent_proficient  , data = attend_race)
summary(attpro)
```

The estimated y-intercept is 46.57 which tells us the predicted value of proficiency when school attendance is zero.

# Discussions

\newpage
# References


