---
title: "The Relationship of Student Attendance and Student Proficiency in English, Language, and Arts "
author: "Tram Anh Hoang, Haitham Anbar, Maria Coronado, Erick Njue"
date: "`r Sys.Date()`"
output: pdf_document
bibliography: references.bib 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, results='hide')


library(here)
library(rio)
library(readr)
library(janitor)
library(tidyverse)
library(knitr)
library(modelsummary)


```


# Abstract
While multiple researchers have suggest a relationship between student attendance and student proficiency, our research examined this relationship using school-level data from the Oregon Department of Education in the latest academic year, 2021-2022.  

Our regression model predicts that any 1 percentage point increase in a school's ratio of regular attenders, there is an additional, statistically significant 0.75 percentage point increase in the school's ratio of students proficient in English, Language, and Arts.  

We also provide a preliminary observation of how the attendance-proficiency relationship varies across student racial and ethnic groups. While our visual exploration suggests that this relationship does vary, additional research is needed to take into account other factors that may correlate with race such as familial background and socio-economic factors.


# Introduction


It is a common assumption that with chronic absenteeism, students’ academic performance decreases, as when students fail to attend school, they miss the opportunity to learn. Previous reviews have stated that chronic absenteeism produces multiple impacts on student outcomes, including lower scores on reading and math, social-emotional development, grade retention and dropout and student discipline [@patnode2018]. On the long run, chronic absenteeism may lead to failure to graduate, underemployment, and financial instability [@henderson2014]. In turn, regular attendance to classes should be a predictor of better performance, and lead to an increased likelihood of graduation. 

Multiple social and economic factors can affect attendance to school. Among these are economic hardship, food insecurity, housing instability, health care access, failing home and school infrastructure, transportation, under-resourced schools, school climate and culture, and attendance culture [@henderson2014]. In fact, school attendance can vary by student group. A national report found variations in attendance and chronic absenteeism when comparing gender, race/ethnicity and language status, free or reduced- price lunch eligibility (proxy for poverty status), and individualized education program status (proxy for disability status) [@garcia2018]. 

School profiles and reports to the Department of Education include relevant information on school characteristics that can be used to understand how social and economic factors affect student performance [@ODEreport].In the state of Oregon, previous data has shown how factors related to absenteeism affect student performance. For example, a report from 2014 showed that in Oregon children of color were disproportionately affected by the consequences of chronic absenteeism. In addition, it showed that for 2009-2010, the student groups with higher absenteeism rates were Black, American Indian and Economically Disadvantaged students [@henderson2014].

Regarding attendance data, in Oregon students are considered "Regular Attenders" if they attended more than 90% of their enrolled days between the beginning of the school year and the first school day in May. In contrast, they are considered chronically absent if they attended 90% or fewer of their enrolled days in this period [@ODEattendance]. 

To explore if regular attendance to classes affected student performance during the school year 2021-2022, we used data from the Oregon Department of Education (ODE) to examine the following research questions:

* Does attendance rate affect student proficiency in English, Language, and Arts? and;
* Does such a relationship between attendance and proficiency vary for different racial and ethnic groups?  


# Methods

We obtained two data sets from the Oregon Department of Education (ODE): attendance rate by school and student group, and proficiency in English, Language, and Arts by student group in the 2021-2022 academic year. We explored the data visually and ran a regression model to examine the relationship between attendance and proficiency in Oregon schools. 

Both data sets we have contain school-level data, broken down by racial and ethnic groups. While the relationship between attendance and student proficiency would be better explained from student-level data, we were not able to obtain data in such format. Our analysis is based on data at the school or institution level. 


## Methodological approach

The research questions and methodological approach we used to answer each of them are:

### RQ1: Does attendance rate affect student proficiency in English, Language, and Arts?  

To answer the first research question, we run a linear regression model, using schools' proficiency rate as the dependent variable, and attendance rate as the independent variable. Our regression equation is:


_Percent of proficient students = Bo + B1 (Percent of regular attendance) + e_

Our hypothesis is that schools with higher student attendance have higher rate of student proficiency in English, Language, and Arts.  

### RQ2: Does such a relationship between attendance and proficiency vary for different racial and ethnic groups?  

For the second question, we looked at whether students' racial and ethnic backgrounds had implications on the effect of attendance on student proficiency. In other words, if students have different racial and ethnic backgrounds, would attendance improve proficiency in the same way. From the literature we reviewed, our hypothesis is that schools will see different impacts of attendance on proficiency for different student groups. To answer this question, we only developed an assessment of the data based on multiple graphics, without fitting a statistical model. 


## Data description and preparation

We used [@R] package [@here] to read in the two data sets from Oregon Department of Education (ODE)into the [@R] environment. We use [@janitor] to clean the names. After preparing the data sets, we display them as tables using [@tidyverse].

```{r read_data}

race <- import(here("data", "pagr_schools_ela_tot_raceethnicity_2122.csv")) %>% 
  clean_names() %>%
  as_tibble()

attend <- import(here("data", "regularattenders_report_2122.csv")) %>% 
  clean_names() %>%
  as_tibble()
```

The first data set is student performance in English Language and Arts (ELA) from all schools in the state. The data set provides the number and percentage of students that are proficient in each of the four levels of ELA proficiency, stratified by school, race and ethnicity, and grade level. This data set has `r nrow(race)` rows and `r ncol(race)` columns.  

The second data set is student attendance data from all the schools in Oregon. This data set provides the number and percentage of regular attenders as well as those of chronically absent students, stratified by school and either race and ethnicity or grade level. This data set has `r nrow(attend)` rows and `r ncol(attend)` columns.  

We cleaned the two data sets by removing rows with no data and rows with suppressed data, marked as “- -” and “*” respectively. We use {dplyr} package functions in [@tidyverse] to do so. 

```{r clean_attendance}
#Summarizing to check the existing student groups in the file
attendsum <- attend %>% 
 group_by(student_group) %>% 
  summarise()

#Filtering the dataframe leaving only student groups related to race. 
attend1 <- attend %>% 
   filter(student_group %in% c("White",
                               "Hispanic/Latino",
                               "Black/African American", 
                               "Asian",
                               "American Indian/Alaska Native",
                               "Native Hawaiian/Pacific Islander",
                               "Multi-Racial")) %>%
  mutate(student_group = recode(student_group, "Native Hawaiian/Pacific Islander" = "Pacific Islander"))



#Removing rows without data
#Mutate attendance data to be numeric
attend1 <- attend1%>% 
  filter(students_included != "*") %>%
  mutate(students_included=as.numeric(students_included),
         number_regular_attenders=as.numeric(number_regular_attenders),
         percent_regular_attenders=as.numeric(percent_regular_attenders),
         number_chronically_absent=as.numeric(number_chronically_absent),
         percent_chronically_absent=as.numeric(percent_chronically_absent))
    
# Note: All schools with >95% and <5% regular attenders were eliminated. 
#These rows corresponded with schools where less than 9 students were included. 

```

```{r clean_profifiency}
#Finding the races considered in the file
racesum <- race %>% 
 group_by(student_group) %>% 
  summarise()

#Filtering the dataframe leaving only student groups related to race.
#Filtered again to remove rows without data on proficiency data
#Transformed number of participants to numeric and the filtered to all >= than 10 students. 
race1 <- race %>% 
   select(-grade_level) %>% 
     filter(student_group == "American Indian/Alaskan Native" |
          student_group == "Asian"|
          student_group == "Black/African American" |
          student_group == "Hispanic/Latino" |
          student_group == "Multi-Racial" |
          student_group == "Pacific Islander" |
          student_group == "White") %>% 
 filter(number_proficient != "*" &
        number_proficient != "--" &
        number_of_participants != "-") %>%
 transform(number_of_participants = as.numeric(number_of_participants))

#Transformed all columns to numbers to be able to compute them. The file had them as characters.
race2<- race1 %>% 
  transform(
    number_proficient = as.numeric(number_proficient), 
    percent_proficient_level_3_or_4 = as.numeric(percent_proficient_level_3_or_4),
    number_level_4 = as.numeric(number_level_4), 
    percent_level_4 = as.numeric(percent_level_4), 
    number_level_3 = as.numeric(number_level_3), 
    percent_level_3 = as.numeric(percent_level_3), 
    number_level_2 = as.numeric(number_level_2), 
    percent_level_2 = as.numeric(percent_level_2), 
    number_level_1 = as.numeric(number_level_1),
    percent_level_1 = as.numeric(percent_level_1), 
    participation_rate = as.numeric(participation_rate)
    )

#Grouped data by school_id and student_group, summed the number of participants and number of proficient students, and re-calculate the percentage of proficient students.
race3 <- race2 %>% 
 group_by(district_id,school_id, student_group) %>% 
 summarise(number_proficient = sum(number_proficient),
           number_of_participants=sum(number_of_participants)) %>%
  mutate(percent_proficient = number_proficient/number_of_participants*100)


```

After cleaning, we merged the two data sets by matching schools’ IDs and student race and ethnicity. Rows with attendance data only or proficiency data only will be removed. We used the _left_join_ in the {dplyr} and _drop_na_ in{tidyr}, all contained in [@tidyverse]. 

```{r join}
#Check that 2 data sets have matching student groups
unique(attend1$student_group)
unique(race3$student_group)

#Join
attend_race <- left_join(attend1, race3, by=c("district_id"="district_id", "institution_id"="school_id", "student_group"="student_group")) %>%
  drop_na()

```

```{r number-of-schools}

school_num <- attend_race %>% 
  group_by(institution) %>% 
  summarize() %>%
  count()

```

After cleaning the data, a total of `r school_num` schools were included in the sample, containing information of the percentage of proficient students and the percentage of regular attendance by race or ethnicity in each school. We use this cleaned data set to run a regression model as described in the Methodological approach section.  

From our clean data set, we identified six racial and ethnic groups: Asian, Black African American, Hispanic/Latino, Multi-racial, Pacific Islander and White. Pacific Islander included students identified as having origins in any of the original peoples of Hawaii, Guam, Samoa, or other Pacific Islands and not Hispanic. We examined the relationship between attendance and student proficiency for each racial and ethnic group using a summary table (attained by using _pivot_wider_, _pivot_longer_, and _kable_) and a visual exploration (attained by using _ggplot_). We did not run regressions for the racial and ethnic groups.



# Results

## The relationship between attendance and proficiency at school level

```{r regression}
attpro <- lm(formula = percent_proficient ~ percent_regular_attenders, data = attend_race)
summary(attpro) 

anova(attpro)

```

We ran a simple regression analysis to examine the relationship between the percentage of regular attenders and the percentage of proficient students in Oregon schools. A total of `r school_num` Oregon schools with existing data were included in the sample.

The results indicated that the percent of regular attendance could statistically significantly predict the percent of proficient students B1=0.75, SE(B1)=0.02, t(2237)=32.5, p<.0001. The results explained that regular attendance accounted for 32% of the explained variability in the percent of proficient students, F (1, 2237) = 1056, p < .0001. The regression equation was: 

_Percentage of proficient students = -6.09 + 0.75 (Percentage of regular attendance)_

This equation indicates that for a 1 percentage point increase in a school's ratio of regular attenders, there is a 0.75 percentage point increase in that school's ratio of proficient students in English, Language, and Arts. We created Figure 1 to demonstrate our data analysis using {ggplot2} of [@tidyverse].  

```{r plot1, results='asis', fig.width = 6.5, fig.height = 4}

#Adding plot for attendance vs. proficiency
attend_race %>%
  group_by(institution_id) %>%
  summarize(students_included = sum(students_included),
            number_regular_attenders = sum(number_regular_attenders),
            percent_regular_attenders = number_regular_attenders/students_included *100,
            number_proficient = sum(number_proficient),
            number_of_participants = sum(number_of_participants),
            percent_proficient = number_proficient/number_of_participants *100) %>%
  ggplot(aes(x=percent_regular_attenders, y=percent_proficient)) +
  geom_point(size = 0.3) +
  geom_smooth(method='lm', color="violet") +
  theme_bw() +
  annotate("text",
           label = expression(hat(Y)~`= -6.09 + 0.75*`~ hat(X)),
           x = 25, y = 75,
           size = 4) +
  labs(title="Student Attendance and Proficiency Rate \n in English, Language, and Arts",
       subtitle="Data from Oregon Education Department",
       x="Rate of Regular Attendance (%)",
       y="Percentage of proficient students (%)",
       caption = "Figure 1")

```


```{r}

modelsummary(attpro, stars=T,
             gof_omit = "Adj.|AIC|BIC|RMSE",
             coef_rename = c("percent_proficient" = "Student proficiency in ELA"),
             notes = c("Notes: Cells represent coefficients and standard errors in parentheses."),
             output = here("scripts", "table1.docx"))

```



## The relationship between attendance and proficiency among different student racial and ethnic groups 

_Table 1_ provides a summary of the mean attendance rate and mean proficiency rate for six racial and ethnic groups across Oregon schools. _Figure 2_ and _Figure 3_ provide a visual display of how attendance and proficiency data spread for different groups.
 
```{r summary-table, results='markup'}

#Transforming relevant columns into interger
attend_race <- attend_race %>%
  transform(
    students_included = as.integer(students_included), 
    number_regular_attenders = as.integer(number_regular_attenders), 
    number_chronically_absent = as.integer(number_chronically_absent)) 


#Summary attendance grouped by race (not by institutions)
attend_summary <- attend_race %>%
    group_by(student_group) %>% 
    summarise(students_included = sum(students_included),
            number_regular_attenders = sum(number_regular_attenders),
            number_chronically_absent = sum(number_chronically_absent)) %>% 
  mutate(percent_regular_attenders = round(number_regular_attenders / students_included * 100,2),
         percent_chron_absent = round(number_chronically_absent / students_included * 100,2)) 

attend_summary <- attend_summary  %>%
  select(-number_chronically_absent, -number_regular_attenders, -students_included, -percent_chron_absent) %>% 
  pivot_wider(
    names_from = "student_group",
    values_from = "percent_regular_attenders") %>% 
  clean_names() %>% 
pivot_longer(
  cols = 1:6,
  names_to = "Race",
  values_to = "Regular attenders") 
  


#Summary proficiency by race (not by institutions)
prof_summary <- attend_race %>%
  group_by(student_group) %>% 
  summarise(number_of_participants = sum(number_of_participants),
            number_proficient = sum(number_proficient)) %>% 
  mutate(percent_prof_race = (number_proficient/number_of_participants*100)) %>% 
  mutate(percent_prof_race = round(percent_prof_race, 2)) %>% 
  select(-number_of_participants, -number_proficient) %>% 
 pivot_wider(names_from = "student_group",
 values_from = "percent_prof_race") %>% 
  clean_names() %>% 
  pivot_longer(
      cols = 1:6,
  names_to = "Race",
  values_to = "Percent proficient") 


mixed_table <- left_join(attend_summary, prof_summary)

mixed_table <- mixed_table %>% 
  mutate(Race = recode(Race, 
                       asian = "Asian",
                       hispanic_latino = "Hispanic/Latino",
                       black_african_american = "Black African American",
                       multi_racial = "Multi Racial",
                       pacific_islander = "Pacific Islander",
                       white = "White"
                       ))

table_1 <- mixed_table %>% 
  kable(caption="Summary table",
        col.names=c("Race/ ethnicity",
                    "Percent regular attenders", 
                    "Percent proficient"))
table_1
    
```


```{r boxplots, fig.width = 6.5, fig.height = 3}



plot4 <- attend_race %>%
    ggplot(aes(x = student_group, y = percent_proficient)) +
   geom_boxplot(fill = "skyblue", alpha = 0.5) +
  theme_bw() +
  labs(x = " ",
       y="Proficient students (%)",
       title = "Rate of proficient students by race in Oregon schools",
       caption = "Figure 2") +
  coord_flip()
plot4



plot3 <- attend_race %>%
    ggplot(aes(x = student_group, y = percent_regular_attenders)) +
   geom_boxplot(fill = "skyblue", alpha = 0.5) +
  theme_bw() +
  labs(x = " ",
       y="Regular attendance (%)",
       title = "Rate of regular attendance by race in Oregon schools",
       caption = "Figure 3") +
  coord_flip()
plot3

```

In term of proficiency, there are vast differences in the mean rate of proficient students between different racial and ethnic groups. If we take White as the reference group with 50.98% students proficient in English, Language, and Arts, the Asian group had a higher proficiency rate of 64.18% while the Pacific Islander, Black/African American, and Hispanic/Latino groups had considerably lower proficiency rates of 15.99%, 19.60%, and 26.11%, respectively.  

In term of attendance, we found a similar pattern, although the differences are less profound. The White reference group had 67.28% regular attenders. The Asian group surpassed the White group with 86.21% regular attenders. This ratio for Pacific Islanders, Black/African Americans, and Hispanics/Latinos is 36.18%, 50.61%, and 56.12%, respectively.  

The Multi-racial group had similar rates as the White group in both attendance and student proficiency with slight differences.  

In Figure 4 below, we used _facet_wrap_ function of {ggplot2} plot the rate of regular attendance by percentage of proficient students based on the race.  

```{r plot2}

#Adding plot for attendance vs. proficiency, faceted by race
attend_race %>%
  ggplot(aes(x=percent_regular_attenders, y=percent_proficient,group=student_group)) +
  geom_point(size = 0.2) +
  geom_smooth(method='lm', color="violet") +  
  facet_wrap(~student_group) +
  theme_bw() +
  labs(title="Student Attendance and Proficiency Rate \n in English, Language, and Arts by Different Races",
       subtitle="Data from Oregon Education Department",
       x="Rate of Regular Attendance (%)",
       y="Percentage of proficient students (%)",
       caption = "Figure 4.") 

```

When faceted by students' racial and ethnic groups, we can see a clear change in the slope of the line. Using White students as the reference group, the Asian group has a steeper slope, suggesting 1% increase in attendance rate results in a higher increase in the rate of proficiency for the Asian group than for the White group. Whereas, the Black/African American group, the Hispanic/Latino group, and the Pacific Islander group had flatter lines, suggesting that the positive impact of attendance rate on proficiency rate is weaker for these groups than for their White counterparts.  

Interestingly, all the data points for the Asian group show up after in the far right part of their plot, indicating uniformly high attendance rate for this group of students - above 60% in all the schools we looked at, while the Pacific Islander group had low attendance - below 70% across all schools. 

# Discussion & Conclusions

Based on data of Oregon schools in the 2021-2022 academic year, our model predicts that for every 1 percentage point increase in attendance rate, schools will see an additional 0.75% percentage point in their ratio of students that are proficient in English, Language, and Arts, and the relationship is statistically significant. This agrees with [@henderson2014] whose study shows that regular attendance to classes should be a predictor of better performance. Our model suggest that a reasonable measure schools can use to improve proficiency rate is to encourage students to attend class more regularly. Policies such as offering free school lunch, addressing bullying, and providing teachers with diversity, equity, and inclusion training have the potential to improve attendance, and should be considered as schools attempt to improve student proficiency.  

Our data comparison across student racial and ethnic groups confirms the findings of [@garcia2018] that attendance and chronic absenteeism vary when comparing various groups and socio-economic factors. [@henderson2014] has similar finding that multiple social and economic factors can affect attendance to school which include economic hardship, food insecurity, housing instability, health care access, failing home and school infrastructure, transportation, under-resourced schools, school climate and culture, and attendance culture.  

Our research also looked at how the relationship between attendance and proficiency varies across student racial and ethnic groups. Based on our preliminary observation and our literature review, we suppose that there are other variables that correlate with racial and ethnic groups and it is those omitted variables, not racial and ethnic groups that have impacts on the relationship between attendance and proficiency. For example, students coming from lower-income households that frequently experience food insecurity might have anxiety and stress that affect their performance, even when their attendance rates are the same as the reference group. While our visual exploration suggests the relationship does vary, additional research is needed to confirm this hypothesis and identify the relationship patterns. This leads us to conclude that interpretation this research could be expanded to include other variables such as household incomes, free school lunch, parents' education attainment, and other relevant factors.  

To conclude, our research, based on data from Oregon schools, confirmed the relationship between attendance and student proficiency as many researchers have suggested. While our preliminary data exploration suggest that this relationship varies across student racial and ethnic groups, there need to be additional research that takes into account other family background and socio-economic factors.  




\newpage
# References


