---
title: "The Relationship of Student Attendance and Student Proficiency in English, Language, and Arts "
author: "Tram Anh Hoang, Haitham Anbar, Maria Coronado, Erick Njue"
date: "`r Sys.Date()`"
output: pdf_document
bibliography: references.bib 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, results='hide')


library(here)
library(rio)
library(readr)
library(janitor)
library(tidyverse)
library(knitr)
library(modelsummary)


```


# Abstract

# Introduction


It is a common assumption that with chronic absenteeism, students’ academic performance decreases, as when students fail to attend school, they miss the opportunity to learn. Therefore,. Previous reviews have stated that chronic absenteeism produces multiple impacts on student outcomes, including lower scores on reading and math, social-emotional development, grade retention and dropout and student discipline [@patnode2018]. On the long run, chronic absenteeism may lead to failure to graduate, underemployment, and financial instability [@henderson2014]. In turn, regular attendance to classes should be a predictor of better performance, and lead to an increased likelihood of graduation. 

Multiple social and economic factors can affect attendance to school. Among these are economic hardship, food insecurity, housing instability, health care access, failing home and school infrastructure, transportation, under-resourced schools, school climate and culture, and attendance culture [@henderson2014]. In fact, school attendance can vary by student group. A national report found variations in attendance and chronic absenteeism when comparing gender, race/ethnicity and language status, FRPL (free or reduced- price lunch) eligibility (our proxy for poverty status), and IEP (individualized education program) status (our proxy for disability status) [@garcia2018]. 

School profiles and reports to the Department of Education include relevant information on school characteristics that can be used to understand how social and economic factors affect student performance [@ODEreport].In the state of Oregon, previous data has shown how factors related to absenteeism affect student performance. For example, a report from 2014 showed that in Oregon children of color were disproportionately affected by the consequences of chronic absenteeism. In addition, it showed that for 2009-2010, the student groups with higher absenteeism rates were Black, American Indian and Economically Disadvantaged students [@henderson2014].

Regarding attendance data, in Oregon students are considered "Regular Attenders" if they attended more than 90% of their enrolled days between the beginning of the school year and the first school day in May. In contrast, they are considered chronically absent if they attended 90% or fewer of their enrolled days in this period [@ODEattendance]. 

To explore if regular attendance to classes affected student performance during the school year 2021-2022, we used data from the Oregon Department of Education (ODE) to examine the following research questions:

* Does attendance rate affect student proficiency in English, Language, and Arts? and;
* Does such a relationship between attendance and proficiency vary for different racial and ethnic groups?  


# Methods

We ran a regression model on attendance data and proficiency data of schools in Oregon. We obtained two different data sets from the Oregon Department of Education (ODE): attendance rate by school and student group, and proficiency in English, Language, and Arts by student group. 

## Data description and preparation

We used [@R] package [@here] to read in the two data sets from Oregon Department of Education (ODE)into the [@R] environment. We use [@janitor] to clean the names. After preparing the data sets, we display them as tables using [@tidyverse].

```{r read_data}

race <- import(here("data", "pagr_schools_ela_tot_raceethnicity_2122.csv")) %>% 
  clean_names() %>%
  as_tibble()

attend <- import(here("data", "regularattenders_report_2122.csv")) %>% 
  clean_names() %>%
  as_tibble()
```

The first data set is student performance in English Language and Arts (ELA) from all schools in the state. The data set provides the number and percentage of students that are proficient in each of the four levels of ELA proficiency, stratified by school, race and ethnicity, and grade level. This data set has `r nrow(race)` rows and `r ncol(race)` columns.  

The second data set is student attendance data from all the schools in Oregon. This data set provides the number and percentage of regular attenders as well as those of chronically absent students, stratified by school and either race and ethnicity or grade level. This data set has `r nrow(attend)` rows and `r ncol(attend)` columns.  

We cleaned the two data sets by removing rows with no data and rows with suppressed data, marked as “- -” and “*” respectively. We use {dplyr} package functions in [@tidyverse] to do so. 

```{r clean_attendance}
#Summarizing to check the existing student groups in the file
attendsum <- attend %>% 
 group_by(student_group) %>% 
  summarise()

#Filtering the dataframe leaving only student groups related to race. 
attend1 <- attend %>% 
   filter(student_group %in% c("White",
                               "Hispanic/Latino",
                               "Black/African American", 
                               "Asian",
                               "American Indian/Alaska Native",
                               "Native Hawaiian/Pacific Islander",
                               "Multi-Racial")) %>%
  mutate(student_group = recode(student_group, "Native Hawaiian/Pacific Islander" = "Pacific Islander"))



#Removing rows without data
#Mutate attendance data to be numeric
attend1 <- attend1%>% 
  filter(students_included != "*") %>%
  mutate(students_included=as.numeric(students_included),
         number_regular_attenders=as.numeric(number_regular_attenders),
         percent_regular_attenders=as.numeric(percent_regular_attenders),
         number_chronically_absent=as.numeric(number_chronically_absent),
         percent_chronically_absent=as.numeric(percent_chronically_absent))
    
# Note: All schools with >95% and <5% regular attenders were eliminated. 
#These rows corresponded with schools where less than 9 students were included. 

```

```{r clean_profifiency}
#Finding the races considered in the file
racesum <- race %>% 
 group_by(student_group) %>% 
  summarise()

#Filtering the dataframe leaving only student groups related to race.
#Filtered again to remove rows without data on proficiency data
#Transformed number of participants to numeric and the filtered to all >= than 10 students. 
race1 <- race %>% 
   select(-grade_level) %>% 
     filter(student_group == "American Indian/Alaskan Native" |
          student_group == "Asian"|
          student_group == "Black/African American" |
          student_group == "Hispanic/Latino" |
          student_group == "Multi-Racial" |
          student_group == "Pacific Islander" |
          student_group == "White") %>% 
 filter(number_proficient != "*" &
        number_proficient != "--" &
        number_of_participants != "-") %>%
 transform(number_of_participants = as.numeric(number_of_participants))

#Transformed all columns to numbers to be able to compute them. The file had them as characters.
race2<- race1 %>% 
  transform(
    number_proficient = as.numeric(number_proficient), 
    percent_proficient_level_3_or_4 = as.numeric(percent_proficient_level_3_or_4),
    number_level_4 = as.numeric(number_level_4), 
    percent_level_4 = as.numeric(percent_level_4), 
    number_level_3 = as.numeric(number_level_3), 
    percent_level_3 = as.numeric(percent_level_3), 
    number_level_2 = as.numeric(number_level_2), 
    percent_level_2 = as.numeric(percent_level_2), 
    number_level_1 = as.numeric(number_level_1),
    percent_level_1 = as.numeric(percent_level_1), 
    participation_rate = as.numeric(participation_rate)
    )

#Grouped data by school_id and student_group, summed the number of participants and number of proficient students, and re-calculate the percentage of proficient students.
race3 <- race2 %>% 
 group_by(district_id,school_id, student_group) %>% 
 summarise(number_proficient = sum(number_proficient),
           number_of_participants=sum(number_of_participants)) %>%
  mutate(percent_proficient = number_proficient/number_of_participants*100)


```

After cleaning, we merged the two data sets by matching schools’ IDs and student race and ethnicity. Rows with attendance data only or proficiency data only will be removed. We used the _left_join_ in the {dplyr} and _drop_na_ in{tidyr}, all contained in [@tidyverse].A 

```{r join}
#Check that 2 data sets have matching student groups
unique(attend1$student_group)
unique(race3$student_group)

#Join
attend_race <- left_join(attend1, race3, by=c("district_id"="district_id", "institution_id"="school_id", "student_group"="student_group")) %>%
  drop_na()

```

```{r number-of-schools}

school_num <- attend_race %>% 
  group_by(institution) %>% 
  summarize() %>%
  count()

```


## Methodological approach

The data sets we obtained from the Oregon Department of Education contain school-level data, broken down by racial and ethnic groups. While the relationship between attendance and student proficiency would be better explained from student-level data, we were not able to obtain data in such format. Therefore, all the data presented here is at the school or institution level. 

After cleaning the data, a total of `r school_num` schools were included in the sample, containing information of the percentage of proficient students and the percentage of regular attendance by race or ethnicity. Six races were identified in the data from ODE: Asian, Black African American, Hispanic/latino, Multi Racial, Pacific Islander and White. Pacific Islander included students identified as having origins in any of the original peoples of Hawaii, Guam, Samoa, or other Pacific Islands and not Hispanic. _Table 1_ contains a summary of this information. We used the function _pivot_wider_ by student group in [@tidyverse] package {tidyr} to have columns by race. This is followed by _pivot_longer_ and another _pivot_wider_ to make the table presentable. We use the _kable_ function found in [@Xie2021-ps] to present the results. 

The research questions and methodological approach we used to answer each of them are:

### RQ1: Does attendance rate affect student proficiency in English, Language, and Arts?  

To answer the first research question, we run a linear regression model, using schools' proficiency rate as the dependent variable, and attendance rate as the independent variable. Our regression equation is:


_Percent of proficient students = Bo + B1 (Percent of regular attendance) + e_

Our hypothesis is that schools with higher student attendance have higher rate of student proficiency in English, Language, and Arts.  

### RQ2: Does such a relationship between attendance and proficiency vary for different racial and ethnic groups?  

For the second question, we looked at whether students' racial and ethnic backgrounds had implications on the effect of attendance on student proficiency. In other words, if students have different racial and ethnic backgrounds, would attendance improve proficiency in the same way. From the literature we reviewed, our hypothesis is that schools will see different impacts of attendance on proficiency for different student groups. To answer this question, we only developed an assessment of the data based on multiple graphics, without fitting a statistical model. 

```{r summary-table, results='markup'}

#Transforming relevant columns into interger
attend_race <- attend_race %>%
  transform(
    students_included = as.integer(students_included), 
    number_regular_attenders = as.integer(number_regular_attenders), 
    number_chronically_absent = as.integer(number_chronically_absent)) 


#Summary attendance grouped by race (not by institutions)
attend_summary <- attend_race %>%
    group_by(student_group) %>% 
    summarise(students_included = sum(students_included),
            number_regular_attenders = sum(number_regular_attenders),
            number_chronically_absent = sum(number_chronically_absent)) %>% 
  mutate(percent_regular_attenders = (number_regular_attenders / students_included * 100)) %>% 
  mutate(percent_regular_attenders = round(percent_regular_attenders, 2)) %>% 
  mutate(percent_chron_absent = (number_chronically_absent / students_included * 100)) %>% 
  mutate(percent_chron_absent = round(percent_chron_absent, 2)) 

attend_summary <- attend_summary  %>%
  select(-number_chronically_absent, -number_regular_attenders, -students_included, -percent_chron_absent) %>% 
  pivot_wider(
    names_from = "student_group",
    values_from = "percent_regular_attenders") %>% 
  clean_names() %>% 
pivot_longer(
  cols = 1:6,
  names_to = "Race",
  values_to = "Regular attenders") 
  


#Summary proficiency by race (not by institutions)
prof_summary <- attend_race %>%
  group_by(student_group) %>% 
  summarise(number_of_participants = sum(number_of_participants),
            number_proficient = sum(number_proficient)) %>% 
  mutate(percent_prof_race = (number_proficient/number_of_participants*100)) %>% 
  mutate(percent_prof_race = round(percent_prof_race, 2)) %>% 
  select(-number_of_participants, -number_proficient) %>% 
 pivot_wider(names_from = "student_group",
 values_from = "percent_prof_race") %>% 
  clean_names() %>% 
  pivot_longer(
      cols = 1:6,
  names_to = "Race",
  values_to = "Percent proficient") 


mixed_table <- left_join(attend_summary, prof_summary)

mixed_table <- mixed_table %>% 
  mutate(Race = recode(Race, 
                       asian = "Asian",
                       hispanic_latino = "Hispanic/latino",
                       black_african_american = "Black African American",
                       multi_racial = "Multi Racial",
                       pacific_islander = "Pacific Islander",
                       white = "White"
                       ))

mixed_table %>% 
  kable(caption="Summary table",
        col.names=c("Race/ ethnicity",
                    "Percent Chronically absent", 
                    "Percent proficient"))

    
```

# Results

```{r regression}
attpro <- lm(formula = percent_proficient ~ percent_regular_attenders, data = attend_race)
summary(attpro) 

anova(attpro)

```

We ran a simple regression analysis to examine the relationship between the percentage of regular attenders and the percentage of proficient students in Oregon schools. A total of `r school_num` Oregon schools with existing data were included in the sample.

The results indicated that the percent of regular attendance could statistically significantly predict the percent of proficient students B1=0.75, SE(B1)=0.02, t(2237)=32.5, p<.0001. The results explained that regular attendance accounted for 32% of the explained variability in the percent of proficient students, F (1, 2237) = 1056, p < .0001. The regression equation was: 

_Percentage of proficient students = -6.09 + 0.75 (Percentage of regular attendance)_

To interpret our results, we created scatter plots using {ggplot2} of [@tidyverse]. In Figure 1, we plotted the percentage of proficient students based on the rate of regular attendance. 

```{r plot1, results='asis', fig.width = 6.5, fig.height = 4}

#Adding plot for attendance vs. proficiency
attend_race %>%
  group_by(institution_id) %>%
  summarize(students_included = sum(students_included),
            number_regular_attenders = sum(number_regular_attenders),
            percent_regular_attenders = number_regular_attenders/students_included *100,
            number_proficient = sum(number_proficient),
            number_of_participants = sum(number_of_participants),
            percent_proficient = number_proficient/number_of_participants *100) %>%
  ggplot(aes(x=percent_regular_attenders, y=percent_proficient)) +
  geom_point(size = 0.3) +
  geom_smooth(method='lm', color="violet") +
  theme_bw() +
  annotate("text",
           label = expression(hat(Y)~`= -6.09 + 0.75*`~ hat(X)),
           x = 25, y = 75,
           size = 4) +
  labs(title="Student Attendance and Proficiency Rate \n in English, Language, and Arts",
       subtitle="Data from Oregon Education Department",
       x="Rate of Regular Attendance (%)",
       y="Percentage of proficient students (%)",
       caption = "Figure 1.")

```


```{r}

modelsummary(attpro, stars=T,
             gof_omit = "Adj.|AIC|BIC|RMSE",
             coef_rename = c("percent_proficient" = "Student proficiency in ELA"),
             notes = c("Notes: Cells represent coefficients and standard errors in parentheses."),
             output = here("scripts", "table1.docx"))

```



## Interpret boxplots here 
 

```{r boxplots, fig.width = 6.5, fig.height = 3}



plot4 <- attend_race %>%
    ggplot(aes(x = student_group, y = percent_proficient)) +
   geom_boxplot(fill = "skyblue", alpha = 0.5) +
  theme_bw() +
  labs(x = " ",
       y="Proficient students (%)",
       title = "Rate of proficient students by race in Oregon schools",
       caption = "Figure 2.") +
  coord_flip()
plot4



plot3 <- attend_race %>%
    ggplot(aes(x = student_group, y = percent_regular_attenders)) +
   geom_boxplot(fill = "skyblue", alpha = 0.5) +
  theme_bw() +
  labs(x = " ",
       y="Regular attendance (%)",
       title = "Rate of regular attendance by race in Oregon schools",
       caption = "Figure 3.") +
  coord_flip()
plot3

```


In Figure 4 below, we used _facet_wrap_ function of {ggplot2} plot the rate of regular attendance by percentage of proficient students based on the race. To answer the second research question, we only included a visual representation of the data, but we didn't fit a regression model. Therefore, we evaluated the data only through a visual assessment.

```{r plot2}

#Adding plot for attendance vs. proficiency, faceted by race
attend_race %>%
  ggplot(aes(x=percent_regular_attenders, y=percent_proficient,group=student_group)) +
  geom_point(size = 0.2) +
  geom_smooth(method='lm', color="violet") +  
  facet_wrap(~student_group) +
  theme_bw() +
  labs(title="Student Attendance and Proficiency Rate \n in English, Language, and Arts by Different Races",
       subtitle="Data from Oregon Education Department",
       x="Rate of Regular Attendance (%)",
       y="Percentage of proficient students (%)",
       caption = "Figure 4.") 

```


# Discussion & Conclusions

## Regression and Plot 1
When looking at school-level data, attendance rate and proficiency rate has a positive relationship. Schools with 1% higher attendance rate saw a x% increase in their percentage of proficient students. Our linear model explains ...% of the data point, with the standard error of...

## Regression and Plot 2
When faceted by students' racial and ethnic groups, we can see a clear change in the slope of the line. Using White students as the reference group, the Asian group has a steeper slope, suggesting 1% increase in attendance rate results in a higher increase in the rate of proficiency for the Asian group than for the White group. Whereas, the Black/African American group, the Hispanic/Latino group, and the Pacific Islander group had flatter line, suggesting that the positive impact of attendance rate on proficiency rate is weaker for these groups than for their White counterparts. 

To explain the varying effect of attendance rate on student proficiency across racial and ethnic groups, we suppose that there are other variables that correlate with racial and ethnic groups and it is those omitted variables, not racial and ethnic groups that have impacts on the relationship between attendance and proficiency. For example, students coming from lower-income households that frequently experience food insecurity might have anxiety and stress that affect their performance, even when their attendance rates are the same as the reference group. This interpretation suggests that this research could be expanded to include other variables such as household incomes, free school lunch, and parents' education attainment.  

Interestingly, all the data points for the Asian group show up in the far right part of their plot, indicating high attendance rate for this group of students - above 60% in all the schools we looked at. Contradictingly, the Pacific Islander group had the lowest attendance - below 70% across all schools.  

# Conclusion - (Merge this in the previous section)

Based on data of Oregon schools in the 2021-2022 academic year, our model predicts that for every 1% increase in attendance rate, schools will see an additional 0.75% in their percentage of students that are proficient in English, Language, and Arts, and the relationship is statistically significant. Our model suggest that a reasonable measure schools can use to improve proficiency rate is to encourage students to attend class more regularly. Policies such as offering free school lunch, addressing bullying, and providing teachers with diversity, equity, and inclusion training have the potential to improve attendance, and should be considered as schools attempt to improve student proficiency.  

Our research also preliminary looked at how the relationship between attendance and proficiency varies across student racial and ethnic groups. While our visual exploration suggests the relationship does vary, additional research is needed to confirm this hypothesis and identify the relationship patterns.



\newpage
# References


